---
title: "Final Porject"
output: pdf_document
date: "2023-07-20"
---

# Data Splitting 
```{r}
set.seed(208941)
Dataset_final = read.csv("/Users/ivylin/Desktop/Guns.csv")
len <- length(Dataset_final$prisoners)
ind <- sample(len, size = 110) # around 10%

# Split the data
dat <- Dataset_final
dat <- dat[, !names(dat)=='d']
inference <- dat[ind, ]
train <- dat[-ind, ]

#variables
violent = Dataset_final$violent
prisoners = Dataset_final$prisoners
cauc = Dataset_final$cauc
afam = Dataset_final$afam
male = Dataset_final$male
income = Dataset_final$income
density = Dataset_final$density
population = Dataset_final$population
law = as.factor(Dataset_final$law)
murder = Dataset_final$murder
robbery = Dataset_final$robbery
state = as.factor(Dataset_final$state)
```


# Explanatory Analysis
``` {r, echo=FALSE, fig.align='center'}
par(mfrow=c(3, 3))
plot(dat$violent ~ dat$murder, ylab = "violent", 
     xlab = "murder")
abline(lm(dat$violent ~ dat$murder), col = "red")

plot(dat$violent ~ dat$robbery, ylab = "violent", 
     xlab = "robbery")
abline(lm(dat$violent ~ dat$robbery), col = "red")

plot(dat$violent ~ dat$prisoners, ylab = "violent", 
     xlab = "prisoners")
abline(lm(dat$violent ~ dat$prisoners), col = "red")

plot(dat$violent ~ dat$afam, ylab = "violent", 
     xlab = "afam")
abline(lm(dat$violent ~ dat$afam), col = "red")

plot(dat$violent ~ dat$cauc, ylab = "violent", 
     xlab = "cauc")
abline(lm(dat$violent ~ dat$cauc), col = "red")

plot(dat$violent ~ dat$male, ylab = "violent", 
     xlab = "male")
abline(lm(dat$violent ~ dat$male), col = "red")

plot(dat$violent ~ dat$population, ylab = "violent", 
     xlab = "population")
abline(lm(dat$violent ~ dat$population), col = "red")

plot(dat$violent ~ dat$income, ylab = "violent", 
     xlab = "income")
abline(lm(dat$violent ~ dat$income), col = "red")

plot(dat$violent ~ dat$density, ylab = "violent", 
     xlab = "density")
abline(lm(dat$violent ~ dat$density), col = "red")
```

```{r}
par(mfrow=c(1, 1))
plot(dat$violent ~ dat$density, main ="",
     ylab = "violence rate", 
     xlab = "density")
abline(lm(dat$violent ~ dat$density))
```

```{r}
tm <- lm(dat$violent ~ dat$murder)
plot(tm$residuals ~ tm$fitted.values)
```


# Step 1: Linear models
```{r}
m1 <- lm(violent ~ murder + robbery + prisoners + afam + cauc + male + population + income + density, data = train)
summary(m1)
```

```{r}
library(car)
vif(m1)
```

```{r}
hist(m1$residuals)
```

```{r}
plot(m1$residuals ~ m1$fitted.values)
```

## Removing cauc from Linear Model 

```{r}
m2 <- lm(violent ~ murder + robbery + prisoners + afam + male + population + income + density, data = train)
summary(m2)
```

```{r}
vif(m2)
```

```{r}
hist(m2$residuals)
```

```{r}
plot(m2$residuals ~ m2$fitted.values)
```


## Trying Other Models

```{r}
m3 <- lm(violent ~ prisoners + cauc + male + population + income + density, data = train)
```

```{r}
summary(m3)
```

```{r}
vif(m3)
```

```{r}
hist(m3$residuals)
```


# Try Transformations
```{r}
new_col <- log(dat$density)
dat1 <- dat
dat1$density <- new_col
inference1 <- dat[ind, ]
train1 <- dat[-ind, ]
```

```{r}
m4 <- lm(violent ~ murder + robbery + prisoners + cauc + male + population + income + density, data = train1)
```

```{r}
plot(m4$residuals ~ m4$fitted.values)
```


# Step 2: Ridge
We only use ridge to check the colinearity we got in step 1, at least for now I did not score the ridge model on our test data when doing model comparison. 

```{r}
ridgeregression <- function(formula,data,lambda) {
# formula: R formula, like for lm()
# data: data.frame containing the variables present in formula
# lambda: numeric scalar > 0, penalization parameter
    
# construct the model matrix
# remove the intercept if present
# centre the covariates
    X <- model.matrix(update.formula(formula,~.-1),data)
    X <- sweep(X,2,colMeans(X),'-')
# compute beta
    sizematrix <- crossprod(X) + diag(rep(lambda,ncol(X))) 
    betahat <- solve(sizematrix,crossprod(X,y))
# return a list containing the coefficients, variance matrix
    list(
        coef = c(mean(data$y),betahat),
        vcov = solve(sizematrix) %*% crossprod(X) %*% solve(sizematrix)
    )
}
```

```{r}
y=train$violent
ridgedframe <- data.frame(y=train$violent, 
                          x1=train$murder, 
                          x2=train$robbery, 
                          x3=train$prisoners, 
                          x4=train$afam, 
                          x5=train$cauc, 
                          x6=train$male, 
                          x7=train$population, 
                          x8=train$income, 
                          x9=train$density)
m5 <- ridgeregression(y~x1+x2+x3+x4+x5+x6+x7+x8+x9, 
                      ridgedframe, lambda = exp(5.5))
```

```{r}
m5$coef
```


# Step 3: Additive models

```{r}
regression_spline_coefficients <- function(y,x,lambda) {
  # fit a regression spline with knots at quantiles of x
  # via penalized least squares with penalty param lambda
  
  ## set up the basis
  basis <- mgcv::smoothCon(s(x,bs='bs'),data.frame(x=x))
  S <- basis[[1]]$S[[1]]
  X <- basis[[1]]$X
  XtX <- crossprod(X)
  # solve
  betahat <- solve(XtX + lambda*S,crossprod(X,y))
  betahat
}
```

```{r}
library(mgcv)
y <- train$violent
x <- train$robbery
regression_spline_coefficients(y, x, 0)
```
```{r}
regression_spline_curve <- function(y,x,lambda,newx) {
  # fit the curve to (y,x) and return the estimated
  # curve at the newx, along with standard errors
  
  ## set up the basis
  basis <- mgcv::smoothCon(s(x,bs='bs'),data.frame(x=x))
  S <- basis[[1]]$S[[1]]
  X <- basis[[1]]$X
  XtX <- crossprod(X)
  # coefficients
  betahat <- solve(XtX + lambda*S,crossprod(X,y))
  # standard errors
  badalexbad <- solve(XtX + lambda*S) # BAD (but you can do it)
  varbeta <- badalexbad %*% XtX %*% badalexbad
  # note: haven't discussed how to estimate sigma
  # predictions
  newX <- mgcv::Predict.matrix(basis[[1]],data.frame(x = newx))
  A <- newX %*% badalexbad %*% t(X)
  yhat <- A %*% y
  varyhat <- tcrossprod(A)
  list(fit = yhat,se = sqrt(diag(varyhat)))
}
```

```{r}
mod <- regression_spline_curve(y, x, lambda = 0, newx = x)
```

```{r}
plot(y, x)
points(mod$fit, x, col="red")
```

```{r}
fit_additive_model <- function(formula,data,lambda) {
  # formula: of the form y ~ s(x1,bs='bs') + ... + s(xp,bs='bs'),
  #   suitable for input into mgcv::smoothCon
  # data: data.frame containing the variables in formula
  # lambda: vector of length p containing smoothing parameters to be used
  
  ## parse the formula
  formula_parsed <- mgcv::interpret.gam(formula)
  y <- data[ ,formula_parsed$response]
  smoothterms <- formula_parsed$smooth.spec
  p <- length(smoothterms)
  
  ## construct the smooth bases
  smooth_bases <- lapply(smoothterms,mgcv::smoothCon,
                         data = data,
                         # absorb the constraints
                         absorb.cons = TRUE
  )
  # get the dimensions in a list, for later
  dimlist <- lapply(lapply(lapply(smooth_bases,'[[',1),'[[','X'),ncol)
  # design matrix
  X <- Reduce(cbind,lapply(lapply(smooth_bases,'[[',1),'[[','X'))
  # construct penalties
  S_list <- lapply(lapply(lapply(smooth_bases,'[[',1),'[[','S'),'[[',1)
  for (j in 1:p)
    S_list[[j]] <- lambda[j] * S_list[[j]]
  S <- Matrix::bdiag(S_list)
  
  ## fit the model
  XtX <- crossprod(X)
  # I swear, we will learn how to do this properly soon...
  badmatrix <- solve(XtX + S) # DON'T EVER DO THIS :( :( :( 
  betahat <- badmatrix %*% crossprod(X,y)
  alpha_est <- mean(y)
  H <- X %*% badmatrix %*% t(X)
  yhat <- H %*% y + alpha_est
  sigma_est <- sqrt(sum((yhat - y)^2) / (length(y) - sum(diag(H))))
  betacov <- (sigma_est^2) * badmatrix %*% XtX %*% badmatrix
  
  ## return
  list(
    formula = formula,
    data = data,
    lambda = lambda,
    betahat = betahat,
    betacov = betacov,
    sigma_est = sigma_est,
    X = X,
    S = S,
    smooth_bases = smooth_bases
  )
}
```

```{r}
## an additive model that does not include not interaction
mod <- fit_additive_model(
  formula = y ~ s(x1, bs="bs") + s(x2, bs="bs") + s(x3, bs="bs") + 
                s(x4, bs="bs") + s(x5, bs="bs") + s(x6, bs="bs") + 
                s(x7, bs="bs") + s(x8, bs="bs") + s(x9, bs="bs"),
  data = ridgedframe,
  lambda = exp(c(1, 1, 1, 1, 1, 1, 1, 1, 1))
)
```


## Fitting Additive Models
There are three possible interactions brought by the three covariates we suspect. Then there are 2^3=8 possible scenarios for these three terms to exist or coexist.

```{r, echo=FALSE}
library(mgcv)
mod1 <- gam(formula = y ~ s(x1, bs="bs") + s(x2, bs="bs") + s(x3, bs="bs") + 
                s(x4, bs="bs") + s(x6, bs="bs") + 
                s(x7, bs="bs") + s(x8, bs="bs") + s(x9, bs="bs"), data = ridgedframe)

#data set with murder x robbery interaction
data2 <- data.frame(y=train$violent, 
                          x1=train$murder, 
                          x2=train$robbery, 
                          x3=train$prisoners, 
                          x4=train$afam, 
                          x5=train$cauc, 
                          x6=train$male, 
                          x7=train$population, 
                          x8=train$income, 
                          x9=train$density,
                          x10=(train$murder)*(train$robbery))

mod2 <- gam(formula = y ~ s(x1, bs="bs") + s(x2, bs="bs") + s(x3, bs="bs") + 
                s(x4, bs="bs") + s(x6, bs="bs") + 
                s(x7, bs="bs") + s(x8, bs="bs") + s(x9, bs="bs") +
                s(x10, bs="bs"), data = data2)              
              
```

```{r, echo=FALSE}
#data set with prisoners x murder interaction
data3 <- data.frame(y=train$violent, 
                          x1=train$murder, 
                          x2=train$robbery, 
                          x3=train$prisoners, 
                          x4=train$afam, 
                          x5=train$cauc, 
                          x6=train$male, 
                          x7=train$population, 
                          x8=train$income, 
                          x9=train$density,
                          x10=(train$murder)*(train$prisoners))

mod3 <- gam(formula = y ~ s(x1, bs="bs") + s(x2, bs="bs") + s(x3, bs="bs") + 
                s(x4, bs="bs") + s(x6, bs="bs") + 
                s(x7, bs="bs") + s(x8, bs="bs") + s(x9, bs="bs") +
                s(x10, bs="bs"), data = data3)
```

```{r, echo=FALSE}
#data set with prisoners x robbery interaction
data4 <- data.frame(y=train$violent, 
                          x1=train$murder, 
                          x2=train$robbery, 
                          x3=train$prisoners, 
                          x4=train$afam, 
                          x5=train$cauc, 
                          x6=train$male, 
                          x7=train$population, 
                          x8=train$income, 
                          x9=train$density,
                          x10=(train$robbery)*(train$prisoners))

mod4 <- gam(formula = y ~ s(x1, bs="bs") + s(x2, bs="bs") + s(x3, bs="bs") + 
                s(x4, bs="bs") + s(x6, bs="bs") + 
                s(x7, bs="bs") + s(x8, bs="bs") + s(x9, bs="bs") +
                s(x10, bs="bs"), data = data4)
```

```{r, echo=FALSE}
#data set with all 3 interactions
data5 <- data.frame(y=train$violent, 
                          x1=train$murder, 
                          x2=train$robbery, 
                          x3=train$prisoners, 
                          x4=train$afam, 
                          x5=train$cauc, 
                          x6=train$male, 
                          x7=train$population, 
                          x8=train$income, 
                          x9=train$density,
                          x10=(train$robbery)*(train$prisoners), 
                    x11 = (train$robbery)*(train$murder), 
                    x12 = (train$murder)*(train$prisoners))

mod5 <- gam(formula = y ~ s(x1, bs="bs") + s(x2, bs="bs") + s(x3, bs="bs") + 
                s(x4, bs="bs") + s(x6, bs="bs") + 
                s(x7, bs="bs") + s(x8, bs="bs") + s(x9, bs="bs") +
                s(x10, bs="bs") + s(x11, bs="bs")+ s(x12, bs="bs"), data = data5)

mod6 <- gam(formula = y ~ s(x1, bs="bs") + s(x2, bs="bs") + s(x3, bs="bs") + 
                s(x4, bs="bs") + s(x6, bs="bs") + 
                s(x7, bs="bs") + s(x8, bs="bs") + s(x9, bs="bs") +
                s(x10, bs="bs") + s(x11, bs="bs"), data = data5)

mod7 <- gam(formula = y ~ s(x1, bs="bs") + s(x2, bs="bs") + s(x3, bs="bs") + 
                s(x4, bs="bs") + s(x6, bs="bs") + 
                s(x7, bs="bs") + s(x8, bs="bs") + s(x9, bs="bs") +
                s(x10, bs="bs") + s(x12, bs="bs"), data = data5)

mod8 <- gam(formula = y ~ s(x1, bs="bs") + s(x2, bs="bs") + s(x3, bs="bs") + 
                s(x4, bs="bs") + s(x6, bs="bs") + 
                s(x7, bs="bs") + s(x8, bs="bs") + s(x9, bs="bs") +
                s(x11, bs="bs") + s(x12, bs="bs"),data = data5)
```


## Removing DC from Training Set and Fitting Models

```{r, echo=FALSE}
train1 <- train[train$state != "District of Columbia", ]

data6 <- data.frame(y=train1$violent, 
                          x1=train1$murder, 
                          x2=train1$robbery, 
                          x3=train1$prisoners, 
                          x4=train1$afam, 
                          x5=train1$cauc, 
                          x6=train1$male, 
                          x7=train1$population, 
                          x8=train1$income, 
                          x9=train1$density)

mm1 <- gam(formula = y ~ s(x1, bs="bs") + s(x2, bs="bs") + s(x3, bs="bs") + 
                s(x4, bs="bs") + s(x6, bs="bs") + 
                s(x7, bs="bs") + s(x8, bs="bs") + s(x9, bs="bs"), data = data6)
```

```{r, echo=FALSE}
data7 <- data.frame(y=train1$violent, 
                          x1=train1$murder, 
                          x2=train1$robbery, 
                          x3=train1$prisoners, 
                          x4=train1$afam, 
                          x5=train1$cauc, 
                          x6=train1$male, 
                          x7=train1$population, 
                          x8=train1$income, 
                          x9=train1$density,
                          x10=(train1$robbery)*(train1$prisoners))

mm2 <- gam(formula = y ~ s(x1, bs="bs") + s(x2, bs="bs") + s(x3, bs="bs") + 
                s(x4, bs="bs") + s(x6, bs="bs") + 
                s(x7, bs="bs") + s(x8, bs="bs") + s(x9, bs="bs") +
                s(x10, bs="bs"), data = data7)
```

```{r, echo=FALSE}
data8 <- data.frame(y=train1$violent, 
                          x1=train1$murder, 
                          x2=train1$robbery, 
                          x3=train1$prisoners, 
                          x4=train1$afam, 
                          x5=train1$cauc, 
                          x6=train1$male, 
                          x7=train1$population, 
                          x8=train1$income, 
                          x9=train1$density,
                          x10=(train1$robbery)*(train1$murder))

mm3 <- gam(formula = y ~ s(x1, bs="bs") + s(x2, bs="bs") + s(x3, bs="bs") + 
                s(x4, bs="bs") + s(x6, bs="bs") + 
                s(x7, bs="bs") + s(x8, bs="bs") + s(x9, bs="bs") +
                s(x10, bs="bs"), data = data8)
```

```{r, echo=FALSE}
data9 <- data.frame(y=train1$violent, 
                          x1=train1$murder, 
                          x2=train1$robbery, 
                          x3=train1$prisoners, 
                          x4=train1$afam, 
                          x5=train1$cauc, 
                          x6=train1$male, 
                          x7=train1$population, 
                          x8=train1$income, 
                          x9=train1$density,
                          x10=(train1$murder)*(train1$prisoners))

mm4 <- gam(formula = y ~ s(x1, bs="bs") + s(x2, bs="bs") + s(x3, bs="bs") + 
                s(x4, bs="bs") + s(x6, bs="bs") + 
                s(x7, bs="bs") + s(x8, bs="bs") + s(x9, bs="bs") +
                s(x10, bs="bs"), data = data9)
```

```{r, echo=FALSE}
# data set with all 3 interactions with DC removed from training set
data10 <- data.frame(y=train1$violent, 
                     x1=train1$murder, 
                     x2=train1$robbery, 
                     x3=train1$prisoners, 
                     x4=train1$afam, 
                     x5=train1$cauc, 
                     x6=train1$male, 
                     x7=train1$population, 
                     x8=train1$income, 
                     x9=train1$density,
                     x10=(train1$robbery)*(train1$prisoners), 
                     x11 = (train1$robbery)*(train1$murder), 
                     x12 = (train1$murder)*(train1$prisoners))

mm5 <- gam(formula = y ~ s(x1, bs="bs") + s(x2, bs="bs") + s(x3, bs="bs") + 
                s(x4, bs="bs") + s(x6, bs="bs") + 
                s(x7, bs="bs") + s(x8, bs="bs") + s(x9, bs="bs") +
                s(x10, bs="bs") + s(x11, bs="bs"), data = data10)

mm6 <- gam(formula = y ~ s(x1, bs="bs") + s(x2, bs="bs") + s(x3, bs="bs") + 
                s(x4, bs="bs") + s(x6, bs="bs") + 
                s(x7, bs="bs") + s(x8, bs="bs") + s(x9, bs="bs") +
                s(x10, bs="bs") + s(x12, bs="bs"), data = data10)

mm7 <- gam(formula = y ~ s(x1, bs="bs") + s(x2, bs="bs") + s(x3, bs="bs") + 
                s(x4, bs="bs") + s(x6, bs="bs") + 
                s(x7, bs="bs") + s(x8, bs="bs") + s(x9, bs="bs") +
                s(x11, bs="bs") + s(x12, bs="bs"), data = data10)

mm8 <- gam(formula = y ~ s(x1, bs="bs") + s(x2, bs="bs") + s(x3, bs="bs") + 
                s(x4, bs="bs") + s(x6, bs="bs") + 
                s(x7, bs="bs") + s(x8, bs="bs") + s(x9, bs="bs") +
                s(x10, bs="bs") + s(x11, bs="bs") + s(x12, bs="bs"), data = data10)
```


# Scoring 
## Scoring Additive Models
```{r, echo=FALSE}
newdat <- data.frame(y = inference$violent, 
                     x1 = inference$murder, 
                     x2 = inference$robbery, 
                     x3 = inference$prisoners, 
                     x4 = inference$afam, 
                     x5 = inference$cauc, 
                     x6 = inference$male, 
                     x7 = inference$population, 
                     x8 = inference$income, 
                     x9 = inference$density)
```

```{r, echo=FALSE}
# score for model 1
sum((newdat$y -  predict(m1, newdata = inference))^2) / length(newdat$y)

# score for model 2
sum((newdat$y -  predict(m2, newdata = inference))^2) / length(newdat$y)

# score for additive model 1
sum((newdat$y -  predict.gam(mod1, newdata = newdat))^2) / length(newdat$y)
```

```{r, echo=FALSE}
newdat <- data.frame(y = inference$violent, 
                     x1 = inference$murder, 
                     x2 = inference$robbery, 
                     x3 = inference$prisoners, 
                     x4 = inference$afam, 
                     x5 = inference$cauc, 
                     x6 = inference$male, 
                     x7 = inference$population, 
                     x8 = inference$income, 
                     x9 = inference$density, 
                     x10=(inference$murder)*(inference$robbery))

sum((newdat$y -  predict.gam(mod2, newdata = newdat))^2) / length(newdat$y)
```

```{r, echo=FALSE}
newdat <- data.frame(y = inference$violent, 
                     x1 = inference$murder, 
                     x2 = inference$robbery, 
                     x3 = inference$prisoners, 
                     x4 = inference$afam, 
                     x5 = inference$cauc, 
                     x6 = inference$male, 
                     x7 = inference$population, 
                     x8 = inference$income, 
                     x9 = inference$density, 
                     x10=(inference$murder)*(inference$prisoners))

sum((newdat$y -  predict.gam(mod3, newdata = newdat))^2) / length(newdat$y)
```

```{r, echo=FALSE}
newdat <- data.frame(y = inference$violent, 
                     x1 = inference$murder, 
                     x2 = inference$robbery, 
                     x3 = inference$prisoners, 
                     x4 = inference$afam, 
                     x5 = inference$cauc, 
                     x6 = inference$male, 
                     x7 = inference$population, 
                     x8 = inference$income, 
                     x9 = inference$density, 
                     x10=(inference$robbery)*(inference$prisoners))

sum((newdat$y -  predict.gam(mod4, newdata = newdat))^2) / length(newdat$y)
```

```{r, echo=FALSE}
newdat <- data.frame(y = inference$violent, 
                     x1 = inference$murder, 
                     x2 = inference$robbery, 
                     x3 = inference$prisoners, 
                     x4 = inference$afam, 
                     x5 = inference$cauc, 
                     x6 = inference$male, 
                     x7 = inference$population, 
                     x8 = inference$income, 
                     x9 = inference$density, 
                     x10=(inference$robbery)*(inference$prisoners), 
                     x11=(inference$murder)*(inference$robbery), 
                     x12=(inference$murder)*(inference$prisoners))


sum((newdat$y -  predict.gam(mod5, newdata = newdat))^2) / length(newdat$y)
sum((newdat$y -  predict.gam(mod6, newdata = newdat))^2) / length(newdat$y)
sum((newdat$y -  predict.gam(mod7, newdata = newdat))^2) / length(newdat$y)
sum((newdat$y -  predict.gam(mod8, newdata = newdat))^2) / length(newdat$y)
```


## Scoring Additive Models without DC observations

```{r, echo=FALSE}
newdat <- data.frame(y = inference1$violent, 
                     x1 = inference1$murder, 
                     x2 = inference1$robbery, 
                     x3 = inference1$prisoners, 
                     x4 = inference1$afam, 
                     x5 = inference1$cauc, 
                     x6 = inference1$male, 
                     x7 = inference1$population, 
                     x8 = inference1$income, 
                     x9 = inference1$density)

sum((newdat$y -  predict.gam(mm1, newdata = newdat))^2) / length(newdat$y)
```


```{r, echo=FALSE}
newdat <- data.frame(y = inference1$violent, 
                     x1 = inference1$murder, 
                     x2 = inference1$robbery, 
                     x3 = inference1$prisoners, 
                     x4 = inference1$afam, 
                     x5 = inference1$cauc, 
                     x6 = inference1$male, 
                     x7 = inference1$population, 
                     x8 = inference1$income, 
                     x9 = inference1$density, 
                     x10=(inference1$murder)*(inference1$robbery))

sum((newdat$y -  predict.gam(mm2, newdata = newdat))^2) / length(newdat$y)
```

```{r, echo=FALSE}
newdat <- data.frame(y = inference1$violent, 
                     x1 = inference1$murder, 
                     x2 = inference1$robbery, 
                     x3 = inference1$prisoners, 
                     x4 = inference1$afam, 
                     x5 = inference1$cauc, 
                     x6 = inference1$male, 
                     x7 = inference1$population, 
                     x8 = inference1$income, 
                     x9 = inference1$density, 
                     x10=(inference1$murder)*(inference1$prisoners))

sum((newdat$y -  predict.gam(mm3, newdata = newdat))^2) / length(newdat$y)
```

```{r, echo=FALSE}
newdat <- data.frame(y = inference1$violent, 
                     x1 = inference1$murder, 
                     x2 = inference1$robbery, 
                     x3 = inference1$prisoners, 
                     x4 = inference1$afam, 
                     x5 = inference1$cauc, 
                     x6 = inference1$male, 
                     x7 = inference1$population, 
                     x8 = inference1$income, 
                     x9 = inference1$density, 
                     x10=(inference1$robbery)*(inference1$prisoners))

sum((newdat$y -  predict.gam(mm4, newdata = newdat))^2) / length(newdat$y)
```

```{r, echo=FALSE}
newdat <- data.frame(y = inference1$violent, 
                     x1 = inference1$murder, 
                     x2 = inference1$robbery, 
                     x3 = inference1$prisoners, 
                     x4 = inference1$afam, 
                     x5 = inference1$cauc, 
                     x6 = inference1$male, 
                     x7 = inference1$population, 
                     x8 = inference1$income, 
                     x9 = inference1$density, 
                     x10=(inference1$robbery)*(inference1$prisoners), 
                     x11=(inference1$murder)*(inference1$robbery), 
                     x12=(inference1$murder)*(inference1$prisoners))


sum((newdat$y -  predict.gam(mm5, newdata = newdat))^2) / length(newdat$y)
sum((newdat$y -  predict.gam(mm6, newdata = newdat))^2) / length(newdat$y)
sum((newdat$y -  predict.gam(mm7, newdata = newdat))^2) / length(newdat$y)
sum((newdat$y -  predict.gam(mm8, newdata = newdat))^2) / length(newdat$y)
```

## Standardizing the Coefficients and Evaluating Level of Influence
```{r, echo=FALSE}
dat55 <- data.frame(y = data5$y, 
                   x1 = (data5$x1 - mean(data5$x1)) / sd(data5$x1), 
                   x2 = (data5$x2 - mean(data5$x2)) / sd(data5$x2), 
                   x3 = (data5$x3 - mean(data5$x3)) / sd(data5$x3), 
                   x4 = (data5$x4 - mean(data5$x4)) / sd(data5$x4), 
                   x5 = (data5$x5 - mean(data5$x5)) / sd(data5$x5), 
                   x6 = (data5$x6 - mean(data5$x6)) / sd(data5$x6), 
                   x7 = (data5$x7 - mean(data5$x7)) / sd(data5$x7), 
                   x8 = (data5$x8 - mean(data5$x8)) / sd(data5$x8), 
                   x9 = (data5$x9 - mean(data5$x9)) / sd(data5$x9), 
                   x10 = (data5$x10 - mean(data5$x10)) / sd(data5$x10), 
                   x11 = (data5$x11 - mean(data5$x11)) / sd(data5$x11), 
                   x12 = (data5$x12 - mean(data5$x12)) / sd(data5$x12))

mod7prime <- gam(formula = y ~ s(x1, bs="bs") + s(x2, bs="bs") + s(x3, bs="bs") + 
                s(x4, bs="bs") + s(x6, bs="bs") + 
                s(x7, bs="bs") + s(x8, bs="bs") + s(x9, bs="bs") +
                s(x10, bs="bs") + s(x12, bs="bs"), data = dat55)
```



``` {r}
#level of influence on all covariates
testpt <- data.frame(x1=c(1), 
                     x2=c(1), 
                     x3=c(1), 
                     x4=c(1), 
                     x5=c(1), 
                     x6=c(1), 
                     x7=c(1), 
                     x8=c(1),
                     x9=c(1),
                     x10=c(1), 
                     x11=c(1), 
                     x12=c(1))

testpt2 <- data.frame(x1=c(1), 
                      x2=c(1), 
                      x3=c(1), 
                      x4=c(1), 
                      x5=c(1), 
                      x6=c(1), 
                      x7=c(1), 
                      x8=c(1),
                      x9=c(1),
                      x10=c(1), 
                      x11=c(1), 
                      x12=c(1))

lst <- c()
for (i in seq_along(testpt2)) {
  temp <- testpt2[, i]  # Store the current value
  testpt2[, i] <- 2     # Set the i-th value to 2
  lst <- c(lst, predict.gam(mod7prime, newdata = testpt2) - predict.gam(mod7, newdata = testpt))
  testpt2[, i] <- temp  # Reset back to the original value
}

abs(lst) 

# density
# afam 
# robbery x prisoners
```






